---
title: 'Stroke: derive cohort'
author: "K. Fleetwood"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache = FALSE}

# 24 Nov: Updated to include people under 40 and over 100

knitr::opts_chunk$set(echo = FALSE, fig.width = 12, height = 7)

library(knitr)
library(tidyverse)
library(DBI)
library(dbplyr)
library(dplyr)
library(janitor)
library(arsenal)
library(sjmisc) # functions for data transformations in the tidyverse

pwd <- readLines("D:\\PhotonUser\\My Files\\Home Folder\\ccu046_db_token_for_r_kf.txt")

# Functions to help with rounding
r_folder <- "D:\\PhotonUser\\My Files\\Home Folder\\r"
source(file.path(r_folder, "tableby_functions_19Jan24.R"))
source(file.path(r_folder, "output_functions_11Jul24.R"))

## connect to databricks
con <- 
  dbConnect(
    odbc:::odbc(),
    "databricks",
    Password = pwd
  )

## connect to the cohort table (runs very quickly)
cohort_tbl <- 
  tbl(
    con, 
    in_schema(
      "dsa_391419_j3w9t_collab", 
      "ccu046_03_out_cohort_combined")
    )
```

This document describes the derivation of the cohort for CCU046_03, analysis of stroke incidence. 
It also examines missing data within this cohort. 
This output was created using linked data from GDPPR, HES and death records.
All counts are rounded to the nearest 5, and all counts under 10 are suppressed.

# Derive cohort
```{r derive_cohort}

# - In GDPPR
# - Alive at the start of follow-up (1 Nov 2019)
# - Born before the end of follow-up (31 Dec 2023)
gdppr_n <- 
cohort_tbl %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31")
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n <- as.numeric(as.data.frame(gdppr_n)[1,1])

# - Exclude people whose date of death is earlier than their date of birth
gdppr_n_dod_check <-
  cohort_tbl %>%
    filter(
      in_gdppr == 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("2023-12-31"),
      is.na(date_of_death)|(date_of_death >= date_of_birth)
    ) %>%
    summarise(n())
gdppr_n_dod_check <- as.numeric(as.data.frame(gdppr_n_dod_check)[1,1])
n_ex_dod <- gdppr_n - gdppr_n_dod_check

# - Exclude people with a stroke prior to the start of the study period
gdppr_n_wo_strk <-
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01"))
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n_wo_strk <- as.numeric(as.data.frame(gdppr_n_wo_strk)[1,1])
n_ex_strk <- gdppr_n - gdppr_n_wo_strk

# Include people whose date of stroke is compatible with their date of death 
# - This will be the cohort for the age-standardised analyses
gdppr_n_dateok <-
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date)
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n_dateok <- as.numeric(as.data.frame(gdppr_n_dateok)[1,1])
n_ex_datebad <- gdppr_n_wo_strk - gdppr_n_dateok

# For people in the cohort for the age-standardised analyses, calculate the number with a stroke
strk_count_as <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    out_stroke_primary_flag == 1
  ) %>%
  summarise(n())

strk_count_as <- as.numeric(as.data.frame(strk_count_as)[1,1])
strk_per_as <- round(100*strk_count_as/gdppr_n_dateok, 1) 

# Exclude people under 40 at the start of follow-up
# Implies date of birth must be <= 31 Dec 1983
gdppr_n_wo_strk_40 <-
  cohort_tbl %>%
    filter(
      in_gdppr %in% 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("1983-12-31"),
      is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
      is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01"))
    ) %>%
    summarise(n())
gdppr_n_wo_strk_40 <- as.numeric(as.data.frame(gdppr_n_wo_strk_40)[1,1])
n_ex_40 <- gdppr_n_wo_strk - gdppr_n_wo_strk_40

# Count number of people who are 101 or older before the start of the follow-up period
# i.e. date_of_birth > 1918-11-01
gpppr_n_40_100 <- 
  cohort_tbl %>%
    filter(
      in_gdppr %in% 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("1983-12-31"),
      is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
      is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
      date_of_birth > as.Date("1918-11-01")
  ) %>%
  summarise(n())
gpppr_n_40_100 <- as.numeric(as.data.frame(gpppr_n_40_100)[1,1])
n_ex_100 <- gdppr_n_wo_strk_40 - gpppr_n_40_100

strk_count <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    date_of_birth > as.Date("1918-11-01"),
    out_stroke_primary_flag == 1
  ) %>%
  summarise(n())

strk_count <- as.numeric(as.data.frame(strk_count)[1,1])
```

* `r format(rnd5(gdppr_n), digits = 10, big.mark = ",")` people in GDPPR, alive at the start of follow-up (1 Nov 2019) and born by the end of follow-up (31 Dec 2023)
* `r format(rnd5(n_ex_strk), digits = 10, big.mark = ",")` people excluded due to stroke prior to start of follow-up
* `r format(rnd5(gdppr_n_wo_strk), digits = 10, big.mark = ",")` people remaining in the cohort
* `r format(rnd5(n_ex_datebad), digits = 10, big.mark = ",")` people excluded due to date of death prior to date of stroke
* `r format(rnd5(gdppr_n_dateok), digits = 10, big.mark = ",")` included in the cohort for the age-standardised incidence analyses
* Of the people included in the cohort for the age-standardised incidence analyses, `r format(rnd5(strk_count_as), digits = 10, big.mark = ",")` (`r strk_per_as`%) had a stroke during the follow-up period 
* `r format(rnd5(n_ex_40), digits = 10, big.mark = ",")` people excluded because they weren't 40 years of age by the end of follow-up
* `r format(rnd5(n_ex_100), digits = 10, big.mark = ",")` aged older than 100 at the start of the study period
* `r format(rnd5(gpppr_n_40_100), digits = 10, big.mark = ",")` people after excluding those younger than 40 and older than 100

```{r baseline_characteristics_as, results = 'hide'}
# Baseline characteristics for people in the cohort for the age-standardised analyses
# Extract data for people with a stroke
cohort_strk_as <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    out_stroke_primary_flag == 1
  ) %>%
  select(
    date_of_birth, sex, ethnicity_19_group, imd_quintile, date_of_death,
    primary_exp_schizophrenia_date, primary_exp_bipolar_disorder_date, primary_exp_depression_date,
    primary_exp_smi_label, primary_exp_smi_date, out_stroke_primary_date
  ) %>%
  collect()

# Need to correct the SMI label to exclude events after the date of stroke
cohort_strk_as <- 
  cohort_strk_as %>%
  mutate(
    sch_date_strk = 
      case_when(
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    smi_strk = factor(smi_strk, levels = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")),
    smi_strk_date = 
      case_when(
        smi_strk == "Schizophrenia" ~ primary_exp_schizophrenia_date,
        smi_strk == "Bipolar disorder" ~ primary_exp_bipolar_disorder_date,
        smi_strk == "Depression" ~ primary_exp_depression_date,
        smi_strk == "No SMI" ~ NA,
      )
  )

# Prepare baseline characteristics table variables
cohort_strk_as <- 
  cohort_strk_as %>%
  mutate(
    age_at_strk = as.numeric(out_stroke_primary_date - date_of_birth)/365.25,
    sex_lab = 
      case_when(
        sex %in% "F" ~ "Female",
        sex %in% "M" ~ "Male"
      ),
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    ethnic_5 = factor(ethnic_5, levels = c("White", "Black", "South Asian", "Mixed", "Any other ethnic group")),
    imd_quintile_lab = 
      case_when(
        imd_quintile %in% 1 ~ "1 (most deprived)",
        imd_quintile %in% 5 ~ "5 (least deprived)",
        TRUE ~ as.character(imd_quintile)
      ),
    imd_quintile_lab = factor(imd_quintile_lab, levels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)"))
  )

labels(cohort_strk_as)  <- 
  c(age_at_strk = 'Age at stroke (years)',
    sex_lab = 'Sex',
    imd_quintile_lab = 'Area based deprivation (IMD quintile)',
    ethnic_5 = 'Ethncity')

tab_cntrls <- 
  tableby.control(
    test = FALSE,
    numeric.stats = c("meansd"), 
    cat.stats = c("countpct_5"), # , "n_per_miss_5"
    cat.simplify = FALSE
  )

bc_tab_as <- 
  tableby(
    smi_strk ~ age_at_strk + sex_lab + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = cohort_strk_as,
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_as$tables$smi_strk$y$stats <- 5*round(bc_tab_as$tables$smi_strk$y$stats/5)

# Repeat for men
bc_tab_m_as <- 
  tableby(
    smi_strk ~ age_at_strk + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_strk_as, sex == "M"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_m_as$tables$smi_strk$y$stats <- 5*round(bc_tab_m_as$tables$smi_strk$y$stats/5)

# Repeat for women
bc_tab_f_as <- 
  tableby(
    smi_strk ~ age_at_strk + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_strk_as, sex == "F"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           #tableby_sum = 'Total', 
           #medianrange_kf = 'Median [Min, Max]',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_f_as$tables$smi_strk$y$stats <- 5*round(bc_tab_f_as$tables$smi_strk$y$stats/5)
```

```{r baseline_characteristics, results = 'hide'}

# Extract data for people with a stroke
cohort_strk <- 
  cohort_tbl %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    date_of_birth > as.Date("1918-11-01"),
    out_stroke_primary_flag == 1
  ) %>%
  select(
    date_of_birth, sex, ethnicity_19_group, imd_quintile, date_of_death,
    primary_exp_schizophrenia_date, primary_exp_bipolar_disorder_date, primary_exp_depression_date,
    primary_exp_smi_label, primary_exp_smi_date, out_stroke_primary_date
  ) %>%
  collect()

# Need to correct the SMI label to exclude events on or after the date of stroke
cohort_strk <- 
  cohort_strk %>%
  mutate(
    sch_date_strk = 
      case_when(
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    smi_strk = factor(smi_strk, levels = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")),
    smi_strk_date = 
      case_when(
        smi_strk == "Schizophrenia" ~ primary_exp_schizophrenia_date,
        smi_strk == "Bipolar disorder" ~ primary_exp_bipolar_disorder_date,
        smi_strk == "Depression" ~ primary_exp_depression_date,
        smi_strk == "No SMI" ~ NA
      )
  )

# Prepare baseline characteristics table variables
cohort_strk <- 
  cohort_strk %>%
  mutate(
    age_at_strk = as.numeric(out_stroke_primary_date - date_of_birth)/365.25,
    sex_lab = 
      case_when(
        sex %in% "F" ~ "Female",
        sex %in% "M" ~ "Male"
      ),
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    ethnic_5 = factor(ethnic_5, levels = c("White", "Black", "South Asian", "Mixed", "Any other ethnic group")),
    imd_quintile_lab = 
      case_when(
        imd_quintile %in% 1 ~ "1 (most deprived)",
        imd_quintile %in% 5 ~ "5 (least deprived)",
        TRUE ~ as.character(imd_quintile)
      ),
    imd_quintile_lab = factor(imd_quintile_lab, levels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)"))
  )

# Create baseline characteristics table
# Only include stroke at age over 40
cohort_strk_40 <- cohort_strk %>% filter(age_at_strk >= 40)
ex_under_40 <- strk_count - nrow(cohort_strk_40)

# Additionally exclude people with a stroke aged greater than or equal to 101
cohort_gt_100 <- sum(cohort_strk_40$age_at_strk >= 101)
cohort_strk_40 <- cohort_strk_40 %>% filter(age_at_strk < 101)

labels(cohort_strk_40)  <- 
  c(age_at_strk = 'Age at stroke (years)',
    sex_lab = 'Sex',
    imd_quintile_lab = 'Area based deprivation (IMD quintile)',
    ethnic_5 = 'Ethncity')

bc_tab <- 
  tableby(
    smi_strk ~ age_at_strk + sex_lab + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = cohort_strk_40,
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           #tableby_sum = 'Total', 
           #medianrange_kf = 'Median [Min, Max]',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab$tables$smi_strk$y$stats <- 5*round(bc_tab$tables$smi_strk$y$stats/5)

# Repeat for men
bc_tab_m <- 
  tableby(
    smi_strk ~ age_at_strk + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_strk_40, sex == "M"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           #tableby_sum = 'Total', 
           #medianrange_kf = 'Median [Min, Max]',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_m$tables$smi_strk$y$stats <- 5*round(bc_tab_m$tables$smi_strk$y$stats/5)

# Repeat for women
bc_tab_f <- 
  tableby(
    smi_strk ~ age_at_strk + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_strk_40, sex == "F"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_f$tables$smi_strk$y$stats <- 5*round(bc_tab_f$tables$smi_strk$y$stats/5)

# Count of people excluded due to a stroke during the follow-up period, before the age of 40
cohort_n <- gpppr_n_40_100 - ex_under_40
strk_per <- round(100*nrow(cohort_strk_40)/cohort_n, 1) 
```
* Of those remaining, `r format(rnd5(ex_under_40), digits = 10, big.mark = ",")` excluded due to stroke before age 40
* `r format(rnd5(cohort_n), digits = 10, big.mark = ",")` people included in cohort
* `r format(rnd5(nrow(cohort_strk_40)), digits = 10, big.mark = ",")` (`r strk_per`%) had a stroke during the follow-up period, aged between 40 and 100 inclusive

# Baseline characteristics of people with a stroke (cohort for age-standardised incidence rates)
```{r by_smi_as, results='asis'}

arsenal:::summary.tableby(
  bc_tab_as, 
  title = "Baseline characteristics of people with a stroke"
)

arsenal:::summary.tableby(
  bc_tab_m_as, 
  title = "Baseline characteristics of men with a stroke"
)

arsenal:::summary.tableby(
  bc_tab_f_as, 
  title = "Baseline characteristics of women with a stroke"
)
```

# Baseline characteristics of people with a stroke
```{r by_smi, results='asis'}

arsenal:::summary.tableby(
  bc_tab, 
  title = "Baseline characteristics of people with a stroke"
)

arsenal:::summary.tableby(
  bc_tab_m, 
  title = "Baseline characteristics of men with a stroke"
)

arsenal:::summary.tableby(
  bc_tab_f, 
  title = "Baseline characteristics of women with a stroke"
)
```

# Explore missing data
```{r miss_explore, results = 'hide', warning = FALSE, message = FALSE}

# Summarise missing ethnicity and deprivation by SMI and stroke status -------------

# Select cohort (exclude people with a stroke before their 40th birthday)
# Redefine SMI
# Flag missing ethnicity and deprivation
# Summarise variables
cohort_miss_sum <-
cohort_tbl %>%
  mutate(
    age_at_strk  = DATEDIFF(sql("DAY"), date_of_birth, out_stroke_primary_date),
    age_at_strk = age_at_strk/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(age_at_strk)|age_at_strk >= 40
  ) %>%
  mutate(
    sch_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    strk = 
      case_when(
        out_stroke_primary_flag %in% 1 ~ "Stroke",
        is.na(out_stroke_primary_flag) ~ "No stroke"
      ),
    miss_ethnicity = as.numeric(is.na(ethnicity_19_group)),
    miss_imd = as.numeric(is.na(imd_quintile)),
    miss_eth_imd = as.numeric(is.na(ethnicity_19_group)|is.na(imd_quintile))
  ) %>%
  group_by(smi_strk, strk) %>%
  summarise(n = n(), miss_eth = sum(miss_ethnicity), miss_dep = sum(miss_imd), miss_any = sum(miss_eth_imd)) %>%
  collect()

smi_lvls = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")
# Format table for output
cohort_miss_sum_out <- 
  cohort_miss_sum %>%
  mutate(
    smi_strk = factor(smi_strk, levels = smi_lvls),
    # flag values less than 10 (only occurs for ethnicity)
    miss_eth_flag10 = miss_eth < 10,
    # round values in accordance with output rules
    n_rnd = 5*round(n/5),
    miss_eth_rnd = ifelse(miss_eth_flag10, 10, 5*round(miss_eth/5)),
    miss_dep_rnd = 5*round(miss_dep/5),
    miss_any_rnd = 5*round(miss_any/5),
    # calculate percentages
    miss_eth_per = format(round(100*miss_eth_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    miss_dep_per = format(round(100*miss_dep_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    miss_any_per = format(round(100*miss_any_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    # create output column
    miss_eth_out = paste0(ifelse(miss_eth_flag10, "<10", miss_eth_rnd), " (", ifelse(miss_eth_flag10, paste0("<", miss_eth_per),miss_eth_per), "%)"),
    miss_dep_out = paste0(miss_dep_rnd, " (", miss_dep_per, "%)"),
    miss_any_out = paste0(miss_any_rnd, " (", miss_any_per, "%)"),
  ) %>%
  arrange(strk, smi_strk)
```

```{r miss_explore_eth, results = 'hide'}
# Baseline characteristics by missingness of ethnicity
# Present: age, sex, deprivation, smi status, stroke status
cohort_miss_by_eth <-
  cohort_tbl %>%
    mutate(
    age_at_strk  = DATEDIFF(sql("DAY"), date_of_birth, out_stroke_primary_date),
    age_at_strk = age_at_strk/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(age_at_strk)|age_at_strk >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    strk = 
      case_when(
        out_stroke_primary_flag %in% 1 ~ "Stroke",
        is.na(out_stroke_primary_flag) ~ "No stroke"
      ),
    miss_ethnicity = ifelse(is.na(ethnicity_19_group), "Ethnicity missing", "Ethnicity available")
  ) %>%
  group_by(miss_ethnicity) %>%
  summarise(
    n          = n(),
    age_mean   = mean(age_at_strk),
    age_sd     = sd(age_at_strk),
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    imd_1      = sum(as.numeric(imd_quintile %in% 1)),
    imd_2      = sum(as.numeric(imd_quintile %in% 2)), 
    imd_3      = sum(as.numeric(imd_quintile %in% 3)),
    imd_4      = sum(as.numeric(imd_quintile %in% 4)),
    imd_5      = sum(as.numeric(imd_quintile %in% 5)),
    imd_miss   = sum(as.numeric(is.na(imd_quintile))),
    smi_sch    = sum(as.numeric(smi_strk %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_strk %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_strk %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_strk %in% "No SMI")),
    strk_yes   = sum(as.numeric(strk %in% "Stroke")),
    strk_no    = sum(as.numeric(strk %in% "No stroke"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_eth <- 
  cohort_miss_by_eth %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    imd_1_fmt      = n_per_fmt(imd_1, n),
    imd_2_fmt      = n_per_fmt(imd_2, n),
    imd_3_fmt      = n_per_fmt(imd_3, n),
    imd_4_fmt      = n_per_fmt(imd_4, n),
    imd_5_fmt      = n_per_fmt(imd_5, n),
    imd_miss_fmt   = n_per_fmt(imd_miss, n),
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    strk_yes_fmt   = n_per_fmt(strk_yes, n),
    strk_no_fmt    = n_per_fmt(strk_no, n)
  )
    
# pivot_longer
cohort_miss_by_eth_out <- 
  cohort_miss_by_eth %>%
  select(miss_ethnicity, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('Ethnicity available', 'Ethnicity missing')

rownames(cohort_miss_by_eth_out) <- 
  c("N", "Age at stroke (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "Area based deprivation (IMD quintile), 1 (most deprived)",
    "Area based deprivation (IMD quintile), 2",
    "Area based deprivation (IMD quintile), 3",
    "Area based deprivation (IMD quintile), 4",
    "Area based deprivation (IMD quintile), 5",
    "Area based deprivation (IMD quintile), Missing",
    "SMI, Schizophrenia",
    "SMI, Bipolar disorder",
    "SMI, Depression", 
    "SMI, No SMI", 
    "Stroke, Yes",
    "Stroke, No")
```

```{r miss_explore_dep, results = 'hide'}
# Baseline characteristics by missingness of deprivation
# Present: age, sex, ethnicity, smi status, stroke status
cohort_miss_by_dep <-
  cohort_tbl %>%
    mutate(
    age_at_strk  = DATEDIFF(sql("DAY"), date_of_birth, out_stroke_primary_date),
    age_at_strk = age_at_strk/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(age_at_strk)|age_at_strk >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    strk = 
      case_when(
        out_stroke_primary_flag %in% 1 ~ "Stroke",
        is.na(out_stroke_primary_flag) ~ "No stroke"
      ),
    miss_imd = ifelse(is.na(imd_quintile), "IMD quintile missing", "IMD quintile available")
  ) %>%
  group_by(miss_imd) %>%
  summarise(
    n          = n(),
    ##
    age_mean   = mean(age_at_strk),
    age_sd     = sd(age_at_strk),
    ##
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    ##
    eth_white       = sum(as.numeric(ethnic_5 %in% "White")),
    eth_mixed       = sum(as.numeric(ethnic_5 %in% "Mixed")),
    eth_south_asian = sum(as.numeric(ethnic_5 %in% "South Asian")),
    eth_black       = sum(as.numeric(ethnic_5 %in% "Black")),
    eth_other       = sum(as.numeric(ethnic_5 %in% "Any other ethnic group")),
    eth_miss        = sum(as.numeric(is.na(ethnic_5))),    
    ##
    smi_sch    = sum(as.numeric(smi_strk %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_strk %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_strk %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_strk %in% "No SMI")),
    ##
    strk_yes   = sum(as.numeric(strk %in% "Stroke")),
    strk_no    = sum(as.numeric(strk %in% "No stroke"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_dep <- 
  cohort_miss_by_dep %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    ##
    eth_white_fmt       = n_per_fmt(eth_white, n),
    eth_mixed_fmt       = n_per_fmt(eth_mixed, n),
    eth_south_asian_fmt = n_per_fmt(eth_south_asian, n),
    eth_black_fmt       = n_per_fmt(eth_black, n),
    eth_other_fmt       = n_per_fmt(eth_other, n),
    eth_miss_fmt        = n_per_fmt(eth_miss, n),
    ##
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    strk_yes_fmt     = n_per_fmt(strk_yes, n),
    strk_no_fmt      = n_per_fmt(strk_no, n)
  )
    
# pivot_longer
cohort_miss_by_dep_out <- 
  cohort_miss_by_dep %>%
  select(miss_imd, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('IMD quintile available', 'IMD quintile missing')

rownames(cohort_miss_by_dep_out) <- 
  c("N", "Age at stroke (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "Ethnicity, White", "Ethnicity, Mixed", "Ethnicity, South Asian", 
    "Ethnicity, Black", "Ethnicity, Any other ethnic group", "Ethnicity, missing",
    "SMI, Schizophrenia", "SMI, Bipolar disorder", "SMI, Depression", "SMI, No SMI", 
    "Stroke, Yes", "Stroke, No")
```

```{r miss_explore_any, results = 'hide'}
# Baseline characteristics by missingnesss either ethnicity and deprivation
# Present age, sex, smi status, stroke status
cohort_miss_by_any <-
  cohort_tbl %>%
    mutate(
    age_at_strk  = DATEDIFF(sql("DAY"), date_of_birth, out_stroke_primary_date),
    age_at_strk = age_at_strk/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_stroke_primary_date)|(out_stroke_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_stroke_primary_date)|(date_of_death >= out_stroke_primary_date),
    is.na(age_at_strk)|age_at_strk >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_stroke_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_stroke_primary_date ~ NA
      ),
    bd_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_stroke_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_stroke_primary_date ~ NA
      ),
    dep_date_strk = 
      case_when(
        is.na(out_stroke_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_stroke_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_stroke_primary_date ~ NA
      ),
    smi_strk = 
      case_when(
        !is.na(sch_date_strk) ~ "Schizophrenia",
        !is.na(bd_date_strk) ~ "Bipolar disorder",
        !is.na(dep_date_strk) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    strk = 
      case_when(
        out_stroke_primary_flag %in% 1 ~ "Stroke",
        is.na(out_stroke_primary_flag) ~ "No stroke"
      ),
    miss_any = ifelse(is.na(ethnicity_19_group)|is.na(imd_quintile), "Missing data", "No missing data")
  ) %>%
  group_by(miss_any) %>%
  summarise(
    n          = n(),
    ##
    age_mean   = mean(age_at_strk),
    age_sd     = sd(age_at_strk),
    ##
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    ##
    smi_sch    = sum(as.numeric(smi_strk %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_strk %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_strk %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_strk %in% "No SMI")),
    ##
    strk_yes     = sum(as.numeric(strk %in% "Stroke")),
    strk_no      = sum(as.numeric(strk %in% "No stroke"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_any <- 
  cohort_miss_by_any %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    ##
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    strk_yes_fmt   = n_per_fmt(strk_yes, n),
    strk_no_fmt    = n_per_fmt(strk_no, n)
  )
    
# pivot_longer
cohort_miss_by_any_out <- 
  cohort_miss_by_any %>%
  select(miss_any, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('No missing data', 'Missing data')

rownames(cohort_miss_by_any_out) <- 
  c("N", "Age at stroke (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "SMI, Schizophrenia", "SMI, Bipolar disorder", "SMI, Depression", "SMI, No SMI",
    "Stroke, Yes", "Stroke, No")
```

```{r miss_out}

kable(
  select(cohort_miss_sum_out, smi_strk, strk, n_rnd, miss_eth_out, miss_dep_out, miss_any_out),
  caption = "Missing data by SMI and stroke status",
  col.names = c("SMI", "Stroke", "N", "Missing ethnicity, n (%)", "Missing deprivation, n (%)", "Missing any, n (%)")
)

kable(cohort_miss_by_eth_out, caption = "Comparison of people in the cohort by ethnicity missingness")
kable(cohort_miss_by_dep_out, caption = "Comparison of people in the cohort by IMD missingness")
kable(cohort_miss_by_any_out, caption = "Comparison of people in the cohort by overall missingness")
```
