---
title: 'Heart failure: derive cohort for analysis'
author: "K. Fleetwood"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache = FALSE}

# 24 Nov: Updated to include people under 40 and over 100

knitr::opts_chunk$set(echo = FALSE, fig.width = 12, height = 7)

library(knitr)
library(tidyverse)
library(DBI)
library(dbplyr)
library(dplyr)
library(janitor)
library(arsenal)
library(sjmisc) # functions for data transformations in the tidyverse

pwd <- readLines("D:\\PhotonUser\\My Files\\Home Folder\\ccu046_db_token_for_r_kf.txt")

# Functions to help with rounding
r_folder <- "D:\\PhotonUser\\My Files\\Home Folder\\r"
source(file.path(r_folder, "tableby_functions_19Jan24.R"))
source(file.path(r_folder, "output_functions_11Jul24.R"))

## connect to databricks
con <- 
  dbConnect(
    odbc:::odbc(),
    "databricks",
    Password = pwd
  )

## connect to the cohort table
cohort_tbl <- 
  tbl(
    con, 
    in_schema(
      "dsa_391419_j3w9t_collab", 
      "ccu046_03_out_cohort_combined")
    )
```

This document describes the derivation of the cohort for CCU046_03, analysis of heart failure incidence. 
It also examines missing data within this cohort. 
This output was created using linked data from GDPPR, HES and death records.
All counts are rounded to the nearest 5, and all counts under 10 are suppressed.

# Derive cohort
```{r derive_cohort}

# - In GDPPR
# - Alive at the start of follow-up (1 Nov 2019)
# - Born before the end of follow-up (31 Dec 2023)
gdppr_n <- 
cohort_tbl %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31")
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n <- as.numeric(as.data.frame(gdppr_n)[1,1])

# - Exclude people whose date of death is earlier than their date of birth
gdppr_n_dod_check <-
  cohort_tbl %>%
    filter(
      in_gdppr == 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("2023-12-31"),
      is.na(date_of_death)|(date_of_death >= date_of_birth)
    ) %>%
    summarise(n())
gdppr_n_dod_check <- as.numeric(as.data.frame(gdppr_n_dod_check)[1,1])
n_ex_dod <- gdppr_n - gdppr_n_dod_check

# - Exclude people with heart failure prior to the start of the study period
gdppr_n_wo_hf <-
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01"))
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n_wo_hf <- as.numeric(as.data.frame(gdppr_n_wo_hf)[1,1])
n_ex_hf <- gdppr_n - gdppr_n_wo_hf

# Include people who's date of heart_failure is compatible with their date of death 
# - This will be the cohort for the age-standardised analyses
gdppr_n_dateok <-
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date)
  ) %>%
  summarise(n()) %>%
  collect()
gdppr_n_dateok <- as.numeric(as.data.frame(gdppr_n_dateok)[1,1])
n_ex_datebad <- gdppr_n_wo_hf - gdppr_n_dateok

# For people in the cohort for the age-standardised analyses, calculate the number with incident heart failure
hf_count_as <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    out_heart_failure_primary_flag == 1
  ) %>%
  summarise(n())

hf_count_as <- as.numeric(as.data.frame(hf_count_as)[1,1])
hf_per_as <- round(100*hf_count_as/gdppr_n_dateok, 1) 

# Exclude people under 40 at the start of follow-up
# Implies date of birth must be <= 31 Dec 1983
gdppr_n_wo_hf_40 <-
  cohort_tbl %>%
    filter(
      in_gdppr %in% 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("1983-12-31"),
      is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
      is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01"))
    ) %>%
    summarise(n())
gdppr_n_wo_hf_40 <- as.numeric(as.data.frame(gdppr_n_wo_hf_40)[1,1])
n_ex_40 <- gdppr_n_wo_hf - gdppr_n_wo_hf_40

# Count number of people who are 101 or older before the start of the follow-up period
# i.e. date_of_birth > 1918-11-01
gpppr_n_40_100 <- 
  cohort_tbl %>%
    filter(
      in_gdppr %in% 1,
      is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
      date_of_birth <= as.Date("1983-12-31"),
      is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
      is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
      date_of_birth > as.Date("1918-11-01")
  ) %>%
  summarise(n())
gpppr_n_40_100 <- as.numeric(as.data.frame(gpppr_n_40_100)[1,1])
n_ex_100 <- gdppr_n_wo_hf_40 - gpppr_n_40_100

hf_count <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    date_of_birth > as.Date("1918-11-01"),
    out_heart_failure_primary_flag == 1
  ) %>%
  summarise(n())

hf_count <- as.numeric(as.data.frame(hf_count)[1,1])
```

* `r format(rnd5(gdppr_n), digits = 10, big.mark = ",")` people in GDPPR, alive at the start of follow-up (1 Nov 2019) and born by the end of follow-up (31 Dec 2023)
* `r format(rnd5(n_ex_hf), digits = 10, big.mark = ",")` people excluded due to heart failure prior to start of follow-up
* `r format(rnd5(gdppr_n_wo_hf), digits = 10, big.mark = ",")` people remaining in the cohort
* `r format(rnd5(n_ex_datebad), digits = 10, big.mark = ",")` people excluded due to date of death prior to date of heart failure
* `r format(rnd5(gdppr_n_dateok), digits = 10, big.mark = ",")` included in the cohort for the age-standardised incidence analyses
* Of the people included in the cohort for the age-standardised incidence analyses, `r format(rnd5(hf_count_as), digits = 10, big.mark = ",")` (`r hf_per_as`%) had heart failure during the follow-up period 
* `r format(rnd5(n_ex_40), digits = 10, big.mark = ",")` people excluded because they weren't 40 years of age by the end of follow-up
* `r format(rnd5(n_ex_100), digits = 10, big.mark = ",")` aged older than 100 at the start of the study period
* `r format(rnd5(gpppr_n_40_100), digits = 10, big.mark = ",")` people after excluding those younger than 40 and older than 100

```{r baseline_characteristics_as, results = 'hide'}

# Baseline characteristics for people in the cohort for the age-standardised analyses

# Extract data for people with heart failure
cohort_hf_as <- 
  cohort_tbl %>%
  filter(
    in_gdppr %in% 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("2023-12-31"),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    out_heart_failure_primary_flag == 1
  ) %>%
  select(
    date_of_birth, sex, ethnicity_19_group, imd_quintile, date_of_death,
    primary_exp_schizophrenia_date, primary_exp_bipolar_disorder_date, primary_exp_depression_date,
    primary_exp_smi_label, primary_exp_smi_date, out_heart_failure_primary_date
  ) %>%
  collect()

# Need to correct the SMI label to exclude events after the date of heart failure
cohort_hf_as <- 
  cohort_hf_as %>%
  mutate(
    sch_date_hf = 
      case_when(
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    smi_hf = factor(smi_hf, levels = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")),
    smi_hf_date = 
      case_when(
        smi_hf == "Schizophrenia" ~ primary_exp_schizophrenia_date,
        smi_hf == "Bipolar disorder" ~ primary_exp_bipolar_disorder_date,
        smi_hf == "Depression" ~ primary_exp_depression_date,
        smi_hf == "No SMI" ~ NA,
      )
  )

# Prepare baseline characteristics table variables
cohort_hf_as <- 
  cohort_hf_as %>%
  mutate(
    age_at_hf = as.numeric(out_heart_failure_primary_date - date_of_birth)/365.25,
    sex_lab = 
      case_when(
        sex %in% "F" ~ "Female",
        sex %in% "M" ~ "Male"
      ),
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    ethnic_5 = factor(ethnic_5, levels = c("White", "Black", "South Asian", "Mixed", "Any other ethnic group")),
    imd_quintile_lab = 
      case_when(
        imd_quintile %in% 1 ~ "1 (most deprived)",
        imd_quintile %in% 5 ~ "5 (least deprived)",
        TRUE ~ as.character(imd_quintile)
      ),
    imd_quintile_lab = factor(imd_quintile_lab, levels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)"))
  )

labels(cohort_hf_as)  <- 
  c(
    age_at_hf = 'Age at heart failure (years)',
    sex_lab = 'Sex',
    imd_quintile_lab = 'Area based deprivation (IMD quintile)',
    ethnic_5 = 'Ethncity'
   )

tab_cntrls <- 
  tableby.control(
    test = FALSE,
    numeric.stats = c("meansd"), 
    cat.stats = c("countpct_5"), # , "n_per_miss_5"
    cat.simplify = FALSE
  )

bc_tab_as <- 
  tableby(
    smi_hf ~ age_at_hf + sex_lab + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = cohort_hf_as,
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_as$tables$smi_hf$y$stats <- 5*round(bc_tab_as$tables$smi_hf$y$stats/5)

# Repeat for men
bc_tab_m_as <- 
  tableby(
    smi_hf ~ age_at_hf + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_hf_as, sex == "M"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_m_as$tables$smi_hf$y$stats <- 5*round(bc_tab_m_as$tables$smi_hf$y$stats/5)

# Repeat for women
bc_tab_f_as <- 
  tableby(
    smi_hf ~ age_at_hf + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_hf_as, sex == "F"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_f_as$tables$smi_hf$y$stats <- 5*round(bc_tab_f_as$tables$smi_hf$y$stats/5)
```

```{r baseline_characteristics, results = 'hide'}

# Extract data for people with heart failure
cohort_hf <- 
  cohort_tbl %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    out_heart_failure_primary_flag == 1
  ) %>%
  select(
    date_of_birth, sex, ethnicity_19_group, imd_quintile, date_of_death,
    primary_exp_schizophrenia_date, primary_exp_bipolar_disorder_date, primary_exp_depression_date,
    primary_exp_smi_label, primary_exp_smi_date, out_heart_failure_primary_date
  ) %>%
  collect()

# Need to correct the SMI label to exclude events after the date of heart failure
cohort_hf <- 
  cohort_hf %>%
  mutate(
    sch_date_hf = 
      case_when(
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    smi_hf = factor(smi_hf, levels = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")),
    smi_hf_date = 
      case_when(
        smi_hf == "Schizophrenia" ~ primary_exp_schizophrenia_date,
        smi_hf == "Bipolar disorder" ~ primary_exp_bipolar_disorder_date,
        smi_hf == "Depression" ~ primary_exp_depression_date,
        smi_hf == "No SMI" ~ NA
      )
  )

# Prepare baseline characteristics table variables
cohort_hf <- 
  cohort_hf %>%
  mutate(
    age_at_hf = as.numeric(out_heart_failure_primary_date - date_of_birth)/365.25,
    sex_lab = 
      case_when(
        sex %in% "F" ~ "Female",
        sex %in% "M" ~ "Male"
      ),
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    ethnic_5 = factor(ethnic_5, levels = c("White", "Black", "South Asian", "Mixed", "Any other ethnic group")),
    imd_quintile_lab = 
      case_when(
        imd_quintile %in% 1 ~ "1 (most deprived)",
        imd_quintile %in% 5 ~ "5 (least deprived)",
        TRUE ~ as.character(imd_quintile)
      ),
    imd_quintile_lab = factor(imd_quintile_lab, levels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)"))
  )

# Create baseline characteristics table
# Only include heart attacks at age over 40
cohort_hf_40 <- cohort_hf %>% filter(age_at_hf >= 40)
ex_under_40 <- hf_count - nrow(cohort_hf_40)

# Additionally exclude people with heart failure aged greater than or equal to 101
cohort_gt_100 <- sum(cohort_hf_40$age_at_hf >= 101)

cohort_hf_40 <- cohort_hf_40 %>% filter(age_at_hf < 101)

labels(cohort_hf_40)  <- 
  c(
    age_at_hf = 'Age at heart failure (years)',
    sex_lab = 'Sex',
    imd_quintile_lab = 'Area based deprivation (IMD quintile)',
    ethnic_5 = 'Ethncity'
   )

tab_cntrls <- 
  tableby.control(
    test = FALSE,
    numeric.stats = c("meansd"), 
    cat.stats = c("countpct_5"), 
    cat.simplify = FALSE
  )

bc_tab <- 
  tableby(
    smi_hf ~ age_at_hf + sex_lab + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = cohort_hf_40,
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab$tables$smi_hf$y$stats <- 5*round(bc_tab$tables$smi_hf$y$stats/5)

# Repeat for men

bc_tab_m <- 
  tableby(
    smi_hf ~ age_at_hf + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_hf_40, sex == "M"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_m$tables$smi_hf$y$stats <- 5*round(bc_tab_m$tables$smi_hf$y$stats/5)

# Repeat for women
bc_tab_f <- 
  tableby(
    smi_hf ~ age_at_hf + includeNA(imd_quintile_lab, first = FALSE) + includeNA(ethnic_5, first = FALSE),
    data = filter(cohort_hf_40, sex == "F"),
    control = tab_cntrls, 
    digits = 1, 
    stats.labels = 
      list(meansd = 'Mean (SD)',
           n_per_miss_5 = 'Missing',
           Nmiss = 'Missing')
  )
bc_tab_f$tables$smi_hf$y$stats <- 5*round(bc_tab_f$tables$smi_hf$y$stats/5)

# Count of people excluded due to heart failure during the follow-up period, before the age of 40
cohort_n <- gpppr_n_40_100 - ex_under_40
hf_per <- round(100*nrow(cohort_hf_40)/cohort_n, 1) 

```
* Of those remaining, `r format(rnd5(ex_under_40), digits = 10, big.mark = ",")` excluded due to heart failure before age 40
* `r format(rnd5(cohort_n), digits = 10, big.mark = ",")` people included in cohort
* `r format(rnd5(nrow(cohort_hf_40)), digits = 10, big.mark = ",")` (`r hf_per`%) had heart failure during the follow-up period, aged between 40 and 100 inclusive

# Baseline characteristics of people with heart failure (cohort for age-standardised incidence rates)
```{r by_smi_as, results='asis'}

arsenal:::summary.tableby(
  bc_tab_as, 
  title = "Baseline characteristics of people with heart failure"
)

arsenal:::summary.tableby(
  bc_tab_m_as, 
  title = "Baseline characteristics of men with heart failure"
)

arsenal:::summary.tableby(
  bc_tab_f_as, 
  title = "Baseline characteristics of women with heart failure"
)
```

# Baseline characteristics of people with heart failure (cohort for Poisson modelling)
```{r by_smi, results='asis'}

arsenal:::summary.tableby(
  bc_tab, 
  title = "Baseline characteristics of people with heart failure"
)

arsenal:::summary.tableby(
  bc_tab_m, 
  title = "Baseline characteristics of men with heart failure"
)

arsenal:::summary.tableby(
  bc_tab_f, 
  title = "Baseline characteristics of women with heart failure"
)
```

# Explore missing data
```{r miss_explore, results = 'hide', warning = FALSE, message = FALSE}

# Summarise missing ethnicity and deprivation by SMI and heart failure status -------------
# Select cohort (exclude people with heart failure before their 40th birthday)
# Redefine SMI
# Flag missing ethnicity and deprivation
# Summarise variables
cohort_miss_sum <-
  cohort_tbl %>%
  mutate(
    age_at_hf  = DATEDIFF(sql("DAY"), date_of_birth, out_heart_failure_primary_date),
    age_at_hf = age_at_hf/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(age_at_hf)|age_at_hf >= 40
  ) %>%
  mutate(
    sch_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    hf = 
      case_when(
        out_heart_failure_primary_flag %in% 1 ~ "Heart failure",
        is.na(out_heart_failure_primary_flag) ~ "No heart failure"
      ),
    miss_ethnicity = as.numeric(is.na(ethnicity_19_group)),
    miss_imd = as.numeric(is.na(imd_quintile)),
    miss_eth_imd = as.numeric(is.na(ethnicity_19_group)|is.na(imd_quintile))
  ) %>%
  group_by(smi_hf, hf) %>%
  summarise(n = n(), miss_eth = sum(miss_ethnicity), miss_dep = sum(miss_imd), miss_any = sum(miss_eth_imd)) %>%
  collect()

cohort_miss_sum

smi_lvls = c("Schizophrenia", "Bipolar disorder", "Depression", "No SMI")
# Format table for output
cohort_miss_sum_out <- 
  cohort_miss_sum %>%
  mutate(
    smi_hf = factor(smi_hf, levels = smi_lvls),
    # flag values less than 10
    miss_eth_flag10 = miss_eth < 10,
    # round values in accordance with output rules
    n_rnd = 5*round(n/5),
    miss_eth_rnd = ifelse(miss_eth_flag10, 10, 5*round(miss_eth/5)),
    miss_dep_rnd = 5*round(miss_dep/5),
    miss_any_rnd = 5*round(miss_any/5),
    # calculate percentages
    miss_eth_per = format(round(100*miss_eth_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    miss_dep_per = format(round(100*miss_dep_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    miss_any_per = format(round(100*miss_any_rnd/n_rnd, digits = 1), digits = 1, nsmall = 1), 
    # create output column
    miss_eth_out = paste0(ifelse(miss_eth_flag10, "<10", miss_eth_rnd), " (", ifelse(miss_eth_flag10, paste0("<", miss_eth_per),miss_eth_per), "%)"),
    miss_dep_out = paste0(miss_dep_rnd, " (", miss_dep_per, "%)"),
    miss_any_out = paste0(miss_any_rnd, " (", miss_any_per, "%)"),
  ) %>%
  arrange(hf, smi_hf)
```

```{r miss_explore_eth, results = 'hide'}

# Baseline characteristics by missingness of ethnicity
# Present: age, sex, deprivation, smi status, heart failure status
cohort_miss_by_eth <-
  cohort_tbl %>%
    mutate(
    age_at_hf  = DATEDIFF(sql("DAY"), date_of_birth, out_heart_failure_primary_date),
    age_at_hf = age_at_hf/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(age_at_hf)|age_at_hf >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    hf = 
      case_when(
        out_heart_failure_primary_flag %in% 1 ~ "Heart failure",
        is.na(out_heart_failure_primary_flag) ~ "No heart failure"
      ),
    miss_ethnicity = ifelse(is.na(ethnicity_19_group), "Ethnicity missing", "Ethnicity available")
  ) %>%
  group_by(miss_ethnicity) %>%
  summarise(
    n          = n(),
    age_mean   = mean(age_at_hf),
    age_sd     = sd(age_at_hf),
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    imd_1      = sum(as.numeric(imd_quintile %in% 1)),
    imd_2      = sum(as.numeric(imd_quintile %in% 2)), 
    imd_3      = sum(as.numeric(imd_quintile %in% 3)),
    imd_4      = sum(as.numeric(imd_quintile %in% 4)),
    imd_5      = sum(as.numeric(imd_quintile %in% 5)),
    imd_miss   = sum(as.numeric(is.na(imd_quintile))),
    smi_sch    = sum(as.numeric(smi_hf %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_hf %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_hf %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_hf %in% "No SMI")),
    hf_yes   = sum(as.numeric(hf %in% "Heart failure")),
    hf_no    = sum(as.numeric(hf %in% "No heart failure"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_eth <- 
  cohort_miss_by_eth %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    imd_1_fmt      = n_per_fmt(imd_1, n),
    imd_2_fmt      = n_per_fmt(imd_2, n),
    imd_3_fmt      = n_per_fmt(imd_3, n),
    imd_4_fmt      = n_per_fmt(imd_4, n),
    imd_5_fmt      = n_per_fmt(imd_5, n),
    imd_miss_fmt   = n_per_fmt(imd_miss, n),
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    hf_yes_fmt   = n_per_fmt(hf_yes, n),
    hf_no_fmt    = n_per_fmt(hf_no, n)
  )
    
# pivot_longer
cohort_miss_by_eth_out <- 
  cohort_miss_by_eth %>%
  select(miss_ethnicity, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('Ethnicity available', 'Ethnicity missing')

rownames(cohort_miss_by_eth_out) <- 
  c("N", "Age at heart failure (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "Area based deprivation (IMD quintile), 1 (most deprived)",
    "Area based deprivation (IMD quintile), 2",
    "Area based deprivation (IMD quintile), 3",
    "Area based deprivation (IMD quintile), 4",
    "Area based deprivation (IMD quintile), 5",
    "Area based deprivation (IMD quintile), Missing",
    "SMI, Schizophrenia", "SMI, Bipolar disorder", "SMI, Depression", "SMI, No SMI", 
    "Heart failure, Yes", "Heart failure, No")
```

```{r miss_explore_dep, results = 'hide'}

# Baseline characteristics by missingness of deprivation
# Present: age, sex, ethnicity, smi status, heart failure status
cohort_miss_by_dep <-
  cohort_tbl %>%
    mutate(
    age_at_hf  = DATEDIFF(sql("DAY"), date_of_birth, out_heart_failure_primary_date),
    age_at_hf = age_at_hf/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(age_at_hf)|age_at_hf >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    hf = 
      case_when(
        out_heart_failure_primary_flag %in% 1 ~ "Heart failure",
        is.na(out_heart_failure_primary_flag) ~ "No heart failure"
      ),
    miss_imd = ifelse(is.na(imd_quintile), "IMD quintile missing", "IMD quintile available")
  ) %>%
  group_by(miss_imd) %>%
  summarise(
    n          = n(),
    age_mean   = mean(age_at_hf),
    age_sd     = sd(age_at_hf),
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    eth_white       = sum(as.numeric(ethnic_5 %in% "White")),
    eth_mixed       = sum(as.numeric(ethnic_5 %in% "Mixed")),
    eth_south_asian = sum(as.numeric(ethnic_5 %in% "South Asian")),
    eth_black       = sum(as.numeric(ethnic_5 %in% "Black")),
    eth_other       = sum(as.numeric(ethnic_5 %in% "Any other ethnic group")),
    eth_miss        = sum(as.numeric(is.na(ethnic_5))),    
    smi_sch    = sum(as.numeric(smi_hf %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_hf %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_hf %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_hf %in% "No SMI")),
    hf_yes   = sum(as.numeric(hf %in% "Heart failure")),
    hf_no    = sum(as.numeric(hf %in% "No heart failure"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_dep <- 
  cohort_miss_by_dep %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    eth_white_fmt       = n_per_fmt(eth_white, n),
    eth_mixed_fmt       = n_per_fmt(eth_mixed, n),
    eth_south_asian_fmt = n_per_fmt(eth_south_asian, n),
    eth_black_fmt       = n_per_fmt(eth_black, n),
    eth_other_fmt       = n_per_fmt(eth_other, n),
    eth_miss_fmt        = n_per_fmt(eth_miss, n),
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    hf_yes_fmt     = n_per_fmt(hf_yes, n),
    hf_no_fmt      = n_per_fmt(hf_no, n)
  )
    
# pivot_longer
cohort_miss_by_dep_out <- 
  cohort_miss_by_dep %>%
  select(miss_imd, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('IMD quintile available', 'IMD quintile missing')

rownames(cohort_miss_by_dep_out) <- 
  c("N", "Age at heart failure (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "Ethnicity, White", "Ethnicity, Mixed", "Ethnicity, South Asian", 
    "Ethnicity, Black", "Ethnicity, Any other ethnic group", "Ethnicity, missing",
    "SMI, Schizophrenia", "SMI, Bipolar disorder", "SMI, Depression", "SMI, No SMI", 
    "Heart failure, Yes", "Heart failure, No")
```

```{r miss_explore_any, results = 'hide'}
# Baseline characteristics by missingnesss either ethnicity and deprivation
# Present age, sex, smi status, heart failure status
cohort_miss_by_any <-
  cohort_tbl %>%
    mutate(
    age_at_hf  = DATEDIFF(sql("DAY"), date_of_birth, out_heart_failure_primary_date),
    age_at_hf = age_at_hf/365.25
  ) %>%
  filter(
    in_gdppr == 1,
    is.na(date_of_death)|(date_of_death > as.Date("2019-11-01")),
    date_of_birth <= as.Date("1983-12-31"),
    date_of_birth > as.Date("1918-11-01"),
    is.na(out_heart_failure_primary_date)|(out_heart_failure_primary_date >= as.Date("2019-11-01")),
    is.na(date_of_death)|is.na(out_heart_failure_primary_date)|(date_of_death >= out_heart_failure_primary_date),
    is.na(age_at_hf)|age_at_hf >= 40
  ) %>%
  mutate(
    ethnic_5 = 
      case_when(
        ethnicity_19_group %in% c("British", "Irish", "Traveller", "Any other White background") ~ "White",
        ethnicity_19_group %in% c("White and Asian", "White and Black African", "White and Black Caribbean", "Any other mixed background") ~ "Mixed",
        ethnicity_19_group %in% c("Bangladeshi", "Indian", "Pakistani") ~ "South Asian",
        ethnicity_19_group %in% c("African", "Caribbean", "Any other Black background") ~ "Black",
        ethnicity_19_group %in% c("Arab", "Chinese", "Any other Asian background", "Any other ethnic group") ~ "Any other ethnic group"
      ),
    sch_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_schizophrenia_date,
        is.na(primary_exp_schizophrenia_date) ~ NA,
        primary_exp_schizophrenia_date < out_heart_failure_primary_date ~ primary_exp_schizophrenia_date,
        primary_exp_schizophrenia_date >= out_heart_failure_primary_date ~ NA
      ),
    bd_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_bipolar_disorder_date,
        is.na(primary_exp_bipolar_disorder_date) ~ NA,
        primary_exp_bipolar_disorder_date < out_heart_failure_primary_date ~ primary_exp_bipolar_disorder_date,
        primary_exp_bipolar_disorder_date >= out_heart_failure_primary_date ~ NA
      ),
    dep_date_hf = 
      case_when(
        is.na(out_heart_failure_primary_date) ~ primary_exp_depression_date,
        is.na(primary_exp_depression_date) ~ NA,
        primary_exp_depression_date < out_heart_failure_primary_date ~ primary_exp_depression_date,
        primary_exp_depression_date >= out_heart_failure_primary_date ~ NA
      ),
    smi_hf = 
      case_when(
        !is.na(sch_date_hf) ~ "Schizophrenia",
        !is.na(bd_date_hf) ~ "Bipolar disorder",
        !is.na(dep_date_hf) ~ "Depression",
        TRUE ~ "No SMI"
      ),
    hf = 
      case_when(
        out_heart_failure_primary_flag %in% 1 ~ "Heart failure",
        is.na(out_heart_failure_primary_flag) ~ "No heart failure"
      ),
    miss_any = ifelse(is.na(ethnicity_19_group)|is.na(imd_quintile), "Missing data", "No missing data")
  ) %>%
  group_by(miss_any) %>%
  summarise(
    n          = n(),
    age_mean   = mean(age_at_hf),
    age_sd     = sd(age_at_hf),
    sex_male   = sum(as.numeric(sex %in% "M")),
    sex_female = sum(as.numeric(sex %in% "F")),
    smi_sch    = sum(as.numeric(smi_hf %in% "Schizophrenia")),
    smi_bd     = sum(as.numeric(smi_hf %in% "Bipolar disorder")),
    smi_dep    = sum(as.numeric(smi_hf %in% "Depression")), 
    smi_no     = sum(as.numeric(smi_hf %in% "No SMI")),
    hf_yes     = sum(as.numeric(hf %in% "Heart failure")),
    hf_no      = sum(as.numeric(hf %in% "No heart failure"))
  ) %>%
  collect()

# Add %s and formatting
cohort_miss_by_any <- 
  cohort_miss_by_any %>%
  mutate(
    age_fmt = paste0(format(age_mean, digits = 1, nsmall = 1), " (", format(age_sd, digits = 1, nsmall = 1), ")"),
    n_rnd = format(rnd5(n), big.mark = ","),
    sex_male_fmt   = n_per_fmt(sex_male, n),
    sex_female_fmt = n_per_fmt(sex_female, n), 
    smi_sch_fmt    = n_per_fmt(smi_sch, n),
    smi_bd_fmt     = n_per_fmt(smi_bd, n),
    smi_dep_fmt    = n_per_fmt(smi_dep, n),
    smi_no_fmt     = n_per_fmt(smi_no, n),
    hf_yes_fmt   = n_per_fmt(hf_yes, n),
    hf_no_fmt    = n_per_fmt(hf_no, n)
  )
    
# pivot_longer
cohort_miss_by_any_out <- 
  cohort_miss_by_any %>%
  select(miss_any, n_rnd, ends_with("fmt")) %>%
  rotate_df(cn = TRUE) %>%
  select('No missing data', 'Missing data')

rownames(cohort_miss_by_any_out) <- 
  c("N", "Age at heart failure (years), [Mean (SD)]", "Sex, Male", "Sex, Female", 
    "SMI, Schizophrenia", "SMI, Bipolar disorder", "SMI, Depression",  "SMI, No SMI", 
    "Heart failure, Yes", "Heart failure, No")
```

```{r miss_out}
kable(
  select(cohort_miss_sum_out, smi_hf, hf, n_rnd, miss_eth_out, miss_dep_out, miss_any_out),
  caption = "Missing data by SMI and heart failure status",
  col.names = c("SMI", "Heart failure", "N", "Missing ethnicity, n (%)", "Missing deprivation, n (%)", "Missing any, n (%)")
)

kable(cohort_miss_by_eth_out, caption = "Comparison of people in the cohort by ethnicity missingness")
kable(cohort_miss_by_dep_out, caption = "Comparison of people in the cohort by IMD missingness")
kable(cohort_miss_by_any_out, caption = "Comparison of people in the cohort by overall missingness")
```
